The development of children's ability to track and predict turn structure in conversation

### 2015-05-01 paper/CasillasFrank-Anticip-Edits.tex

\usepackage{graphicx}
\usepackage{lineno}
\usepackage{amssymb}
\usepackage{wrapfig}
\usepackage{tipa}
\usepackage{float}

@@

\author[MPI]{Marisa Casillas\corref{cor1}}
\address[MPI]{Max Planck Institute for Psycholinguistics, Nijmegen}
%\address[StanfordLX]{Department of Linguistics, Stanford University}
\cortext[cor1]{Corresponding author}

\author[StanfordPSY]{Michael C. Frank}

\address[StanfordPSY]{Department of Psychology, Stanford University}

\begin{abstract}
We investigate language acquisition through the lens of a fundamental conversational skill: turn taking. Children begin developing turn-taking skills in infancy, but take several years to assimilate their growing knowledge of language with this ability. In two eye-tracking experiments with children across a wide developmental range, we measured spontaneous predictions about upcoming speaker change while controlling the amount of linguistic information available. We found that children already predicted upcoming turns at age one, but they integrated linguistic cues differently at different ages. Children under three showed a general advantage for prosody over lexicosyntax, contrary to prior findings for adults. Children two and older showed more predictive switches for questions than non-questions, but only when lexical information was available. We found no evidence that lexicosyntax alone guides turn prediction---instead, participants' performance was best overall with access to lexicosyntax and prosody together. Our results point to the importance of studying children's linguistic processing in the context of conversation.
\end{abstract}

@@

Spontaneous conversation is a universal context for using and learning language. Like other types of human interaction, it is organized at its core by the roles and goals of its participants. But what sets conversation apart is its structure: sequences of interconnected, communicative actions that take place across alternating turns at talk. Sequential, turn-based structures in conversation are strikingly uniform across language communities and linguistic modalities. Turn-taking behaviors are also cross-culturally consistent in their basic features and the details of their implementation \citep{de-vosInPrep, dingemanse2013, stivers2009}. How does this ability develop?

Children participate in sequential coordination with their caregivers starting at three months of age---before they can rely on any linguistic cues in taking turns \citep[see, among others, ][]{bateson1975, hilbrinkInPrep, jaffe2001, snow1977}. Of course, infant turn taking is different from adult turn taking in several ways. Infant turn taking is heavily scaffolded by caregivers, has distinct timing in comparison to adult turn taking, and lacks semantic content \citep{hilbrinkInPrep, jaffe2001}. However, children's early, turn-structured social interactions are presumably a critical precursor to their conversational turn taking. Along these lines, early non-verbal interactions establish the protocol by which children come to use language with others. And as they acquire language, they also come to integrate it into the preverbal turn-taking systems.

In this study, we investigate when children begin to make predictions about upcoming turn structure in conversation, and how they integrate language into their predictions as they grow older. In what follows, we first give a basic review of turn-taking research and the state of current knowledge about adult turn prediction. We then discuss recent work on the development of turn-taking skills before turning to the details of our own study.

\subsection{Turn taking}

Turn taking itself is not unique to conversation. Many other human activities are organized around sequential turns at action. Traffic intersections and computer network communication both use turn-taking systems. Children's early games (e.g., give-and-take, peek-a-boo) have built-in, predictable turn structure \citep{ratner1978, ross1987}. Even monkeys take turns: Non-human primates such as marmosets and Campbell's monkeys vocalize contingently with each other in both natural and lab-controlled environments \citep{lemasson2011, takahashi2013}. In all these cases, turn taking serves as a protocol for interaction, allowing the participants to coordinate with one another through sequences of contingent action.

Conversation distinguishes itself from non-conversational turn-taking behaviors by the complexity of the turn sequencing involved. In the examples above (traffic, games, and monkeys) the set of sequence and action types is far more limited and predictable than what we find in everyday talk. For example, conversational turns come grouped into semantically-contingent sequences of action. The groups can span turn-by-turn exchanges (e.g., simple question--response, ``How are you?''--``Fine.'') or sequence-by-sequence exchanges (e.g., reciprocals, ``How are you?''--``Fine, and you?''--``Great!''). Sequences of action drive the conversation forward into the next, relevant sequences of talk (e.g., ''And you?''--``Great!''--``Why's that?''; \citealp{schegloff2007}). To take a turn, participants need to make predictions about what conversational content will be relevant next. In some cases, relevant next turns are somewhat obvious (e.g., question--response) while, in other cases, there are multiple relevant next actions to choose from, or no obvious next action at all (e.g., after a closed sequence).

Despite this complexity, conversational turn taking is often precise in its timing. Across a diverse sample of conversations in 10 languages, one study found a consistent average transition time of 0--200 msec at points of speaker switch \citep{stivers2009}. Experimental results and current models of speech production suggest that it takes approximately 600 msec to produce a content word, and even longer to produce a simple utterance \citep{griffin2000, levelt1989}. So in order to achieve 200 msec turn transitions, speakers must begin formulating their response before the prior turn has ended \citep{levinson2013}. Moreover, to formulate their response early on, speakers must track and anticipate what types of response might become relevant next. They also need to predict the content and form of upcoming speech so that they can launch their articulation at exactly the right moment. Prediction thus plays a key role in timely turn taking.

\subsection{Adults' turn prediction}

Adults have a lot of information at their disposal to help make accurate predictions about upcoming turn content. Lexical, syntactic, and prosodic information (e.g., \textit{wh}- words, subject-auxiliary inversion, and list intonation) can all inform addressees about upcoming linguistic structure \citep{de-ruiter2006, duncan1972, ford1996, torreiraUndRev}. Non-verbal cues (e.g., gaze, posture, and pointing) often appear at turn-boundaries and can sometimes act as late indicators of an upcoming speaker switch \citep{rossano2009, stivers2010}. Additionally, the sequential context of a turn can make it clear what will come next: answers after questions; thanks or denial after compliments, et cetera \citep{schegloff2007}.

Prior work suggests that adult listeners primarily use lexicosyntactic information to accurately predict upcoming turn structure \citep{de-ruiter2006}. De Ruiter and colleagues \citeyearpar{de-ruiter2006} asked participants to listen to snippets of spontaneous conversation and to press a button whenever they anticipated that the current speaker was about to finish her turn. The speech snippets were controlled for the amount of linguistic information present; some were normal, but others had flattened pitch, low-pass filtered speech, or further manipulations. De Ruiter and colleagues found that, with pitch-flattened speech, the timing of participants' button responses was comparable to their timing with the full linguistic signal. But, when no lexical information was available, participants' responses were significantly earlier than the full linguistic signal. The authors concluded that lexicosyntactic information\footnote{The ``lexicosyntactic'' condition only included flattened pitch and so was not exclusively lexicosyntactic---the speech would still have residual prosodic structure, including syllable duration and intensity.} was necessary and possibly sufficient for turn-end projection, while intonation was neither necessary nor sufficient. Congruent evidence comes from studies varying the predictability of lexicosyntactic and pragmatic content: Adults anticipate turn ends better when they can more accurately predict the exact words that will come next (\citealp{magyari2012}; see also \citealp{magyariUndRev}). They can also identify speech acts within the first word of an utterance \citep{gisladottirUndRev}, allowing them to start planning their response at the first moment possible \citep{bogelsmagyariInPrep}.

The role of prosody for adult turn-prediction is still a matter of debate. De Ruiter and colleagues' (2006) experiment only tested the role of intonation, which is only a partial index of prosody. Prosodic structure is also tied closely to the syntax of an utterance, and so the two linguistic signals are difficult to control independently \citep{ford1996}. \citet*{torreiraUndRev} used a combination of button-press and verbal responses to investigate the relationship between lexicosyntactic and prosodic cues in turn-end prediction. Critically, their stimuli were cross-spliced so that each item had full prosodic cues to accompany the lexicosyntax. Because of the splicing, they were able to create items that had syntactically-complete units with no intonational phrase boundary at the end. Participants never verbally responded or pressed the ``turn-end'' button when hearing a syntactically-complete phrase without an intonational phrase boundary. And when intonational phrase boundaries were embedded in multi-utterance turns, participants were tricked into pressing the ``turn-end'' button 29\% of the time. Their results suggest that listeners actually do rely on prosodic cues to execute a response (see also de \citet{de-ruiter2006}:525). These experimental findings corroborate other corpus and experimental work promoting a combination of cues (lexicosyntactic, prosodic, and pragmatic) as key for accurate turn-end prediction \citep{duncan1972, ford1996, hirvenkari2013}.

In sum, adults accurately and spontaneously make predictions about upcoming turn structure. Their predictions rely on a sophisticated body of knowledge about linguistic structure, non-verbal signals, and social actions. Knowing this, we could expect that children's acquisition of turn-taking skills is closely tied to their knowledge about language, gaze, gesture, and social cues. But children's turn taking starts early in infancy, long before first words or gestures emerge. So a primary role for lexicosyntactic cues doesn't fit well with children's pre-verbal turn taking.

\subsection{Children's turn prediction}

@@

The majority of work on children's early turn taking has focused on observations of spontaneous interaction. Children's first turn-like structures appear as early as two to three months in proto-conversation with their caregivers \citep{bruner1975, bruner1985}. During proto-conversations, caregivers interact with their infants as if they were capable of making meaningful contributions; they take every look, vocalization, arm flail and burp as ``utterances'' in the joint discourse \citep{bateson1975, jaffe2001, snow1977}. Infants catch onto the structure of proto-conversations quickly. By three to four months they notice disturbances to the contingency of their caregivers' response and, in reaction, change the rate and quality of their vocalizations \citep{k-bloom1988, masataka1993}. Infants at this age also notice changes to social contingency outside of turn structure. In the Still Face paradigm, caregivers interact with their infants and then suddenly halt, taking on a neutral expression with a sustained gaze. When faced with this sudden disappearance of social contingency, infants three months and older try a range of methods to reinitiate the interaction, such as vocalization, reaching, and smiling before looking away or getting upset \citep{rochat1998, toda1993}.

The timing of children's responses to their caregivers' speech shows a non-linear pattern of fall-rise-fall from early infancy to middle childhood. A recent study by Hilbrink et al. (in preparation) finds that infants' turn timing at three months is often too early or too late: they start vocalizing in overlap on nearly 40\% of their caregivers' turns, and their non-overlapped vocalizations come after an average inter-turn silent gap of 675 msec (adult average: 200 msec). Between four and nine months, children begin to reduce the number of turns happening in overlap while also improving on their average response latency. But then, later on, children's response latencies slow down again, peaking at average gaps of 1100 msec at nine months, with only very gradual improvement after that \citep{hilbrinkInPrep}. While children's avoidance of overlap is nearly adult-like by nine months, the timing of their non-overlapped responses stays much longer than the 200 msec standard for the next several years \citep{garvey1984, ervin-tripp1979}.

The protracted development of children's timing may be attributable to their linguistic development: Taking turns on time is easier when the response is a simple vocalization rather than a linguistic utterance. Integrating language into the turn-taking system may be one major factor in children's delayed responses \citep{casillasUndRev}. If response planning (i.e., language production) is the primary hurdle in children's spontaneous turn taking, we should find evidence that children understand turn-taking behaviors before they are able to produce the behaviors themselves; this hypothesis has been recently explored in experimental settings.

@@

Children begin to develop specific expectations about conversational behavior before they begin to speak. Sometime between four and six months, children begin to attend differently to face-to-face and back-to-back conversation; six-month-olds follow conversational speakers more with their gaze when at least one speaker is looking at the other \citep{augusti2010}. At ten months, infants expect people to look and talk at other people, and not to objects \citep{beier2012}. At twelve months infants expect to see responses to verbal (but not non-speech) utterances in face-to-face contexts \citep{thorgrimssonUndRev}.

There are mixed results regarding when children begin to anticipate turn structure in conversation. One study found that 12-month-olds make more predictive gaze shifts to a responder while watching human verbal conversation compared to conversation-like interactions with objects \citep{bakker2011}, but another only found a similar effect at 36 months \citep{hofsten2009}. Neither of these two studies had baselines to which the turn-relevant looking behavior could be compared, however. A baseline measurement is critical because there may be developmental differences in gaze shifting between conversational participants, even if the shifting is not related to turn structure. Such developmental differences could produce artifactual changes in measures of turn-contingent shifting.

Keitel and colleagues \citeyearpar{keitel2013} addressed the random baseline issue in their study of 6-, 12-, 24-, and 36-month-olds. They asked participants to watch short videos of conversation, and tracked their eye movements at points of speaker change. They found that children's anticipatory gaze frequency was only greater than chance for 36-month-olds and adults. Their study was the first to focus on the role of linguistic processing in children's turn predictions. They showed their participants two types of conversation videos: one normal and one with flattened pitch (i.e., with flattened intonation contours), finding that only 36-month-olds were affected by a lack of intonation contours. The adult control group made equal numbers of anticipatory looks in the videos, with and without intonation contours, consistent with prior adult findings \citep{de-ruiter2006}. Keitel and colleagues concluded that children's ability to predict upcoming turn structure relies on their ability to comprehend the stimuli (emerging around 36 months), especially with respect to semantic access. They also suggest that intonation takes a secondary role in turn prediction, but only \textit{after} children acquire more sophisticated, adult-like language comprehension systems (sometime after 36 months).

Although the Keitel et al. \citeyearpar{keitel2013} study constitutes a substantial advance over previous work, it nevertheless has several limitations. Because these limitations directly inform our own study design, we review them in some detail. First, their estimates of baseline gaze frequency (``random'' in their terminology) were computed from switches that occurred specifically while the speaker's turn was ongoing; in contrast, estimates of anticipatory shifting frequency were computed from switches that occurred within the time window immediately surrounding inter-turn gaps. Assuming that gaze switches happen most often during the inter-turn gap \citep{hirvenkari2013}, the switches made during ongoing speech (their baseline) would be minimized. This baseline thus maximized the possibility of finding a difference between ``random'' and anticipatory gaze frequencies. A more conservative baseline would be to compare participants' looking behavior at turn transitions to their looking behavior during randomly-selected windows of time throughout the stimulus. We follow this conservative approach in our work.

Second, the conversation stimuli they used were somewhat unusual. The average gap between turns was 900 msec, which is much longer than typical adult timing, where gaps average around 200 msec \citep{stivers2009}. The speakers in the videos were also asked to minimize their movements while performing a scripted and adult-directed conversation, which would have created a somewhat unnatural stimulus. In addition, in order to produce more naturalistic conversation, it would have been ideal to localize sources for the two voices in the video (i.e., to have the voices come out of separate left and right speakers). But both voices were recorded and played back on the same audio channel, which may have made it more difficult to distinguish the two talkers. Again, we attempt to address these issues in our current study.

Despite these minor methodological issues, the Keitel et al. \citeyearpar{keitel2013} study still demonstrates intriguing age-based differences in children's ability to predict upcoming turn structure, and the results suggest that both semantic and intonational development \textit{do} play a role in children's looking patterns. Our current work thus takes this paradigm as our starting point.\footnote{See also \citet{casillas2012, casillas2013}.}

\subsubsection{Prosodic development}
The roles of prosody and lexicosyntax in children's turn predictions are currently unknown, but children understand more about prosody early in life. Children begin to acquire prosody in the womb, and can distinguish their native language's rhythm type from others (e.g., syllable-timed vs. stress-timed) 2--5 days after birth \citep{mehler1988, moon1993, nazzi2003}. Beginning between four and five months, infants prefer pauses in speech to be inserted at prosodic boundaries, and by 6 months they can start using prosodic markers to pick out sub-clausal syntactic units \citep{jusczyk1995, soderstrom2003}. They show preference for the typical stress patterns of their native language over others by 6--9 months (e.g., iambic vs. trochaic), and can use prosodic information to segment the speech stream into smaller chunks from 8 months onward \citep{johnson2001, jusczyk1993, morgan1995}. In comparison, children show only a limited lexical inventory at six months, and begin to recognize function words just before their first birthdays, with syntactic categorization beginning around 14 months \citep{bergelson2013, shi2010}. Two-month-olds notice changes in word order, but this ability appears to rely on prosodic cueing \citep{mandel1996}. Generally speaking then, our current knowledge about children's linguistic development points to a possible early advantage for prosody in children's turn-taking predictions.

\subsection{The Current Study}
We report here on the role of linguistic processing in children's predictions about upcoming turn structure. We focus in particular on how children use prosodic and lexicosyntactic information to make their predictions. Prior work has focused mainly on lexicosyntax and intonation, and not on prosody proper (\citealp{de-ruiter2006}; \citealp{keitel2013}, but see \citealp{torreiraUndRev}), even though infants seem to acquire the basic rhythmic properties of the prosodic signal first \citep{mehler1988, moon1993, nazzi2003}.

In two eye-tracking experiments, we measured children's anticipatory gaze to upcoming responders while controlling for the amount of lexicosyntactic and prosodic information available. Experiment 1 uses a paradigm in which English-speaking participants view video clips of naturalistic conversations from several different languages in order to create conditions under which lexicosyntactic information was differentially available, but prosodic and non-linguistic cues were intact. This paradigm showed minimal differences between the predictive looking behavior of preschoolers and adults. In Experiment 2, we created artificial (puppet) visual stimuli to better separate these linguistic cues from one another. In this more controlled paradigm, we found that children's predictive looking behavior improved from ages one to six, but that even one-year-olds made more anticipatory looks than would be expected by chance.

In both experiments children consistently looked faster to responders after hearing questions, compared to non-questions. Both prosodic and lexicosyntactic information played a role in children's predictions about turn structure, but the two information sources were used differently at different ages. Our findings overall support an account in which predictive processes for turn taking in conversation are present early, but their integration with linguistic information takes substantial practice.

@@

\section{General Discussion}
\label{sec:gendisc}
Children begin to develop conversational turn-taking skills long before their first words \citep{bateson1975, hilbrinkInPrep, jaffe2001, snow1977}. As they acquire language, they also acquire the information needed to make accurate predictions about upcoming turn structure. Until recently, we have had very little data on how children weave language into their already-existing turn-taking behaviors.

@@

Our findings are consistent with observational work on children's turn taking. Young children are slow turn-takers in spontaneous conversation \citep{casillasUndRev} even though they begin taking turns in infancy \citep{hilbrinkInPrep, jaffe2001}. We found that children begin to spontaneously predict upcoming turn structure at age one. Children's gaze switches were also most common in the inter-turn gap, suggesting that they expected an immediate response from the addressee. Our results therefore support the idea of an early-developed competence for the timing of turn taking. But because children do not themselves use adult-like conversational timing until much later \citep[age six; ][]{ervin-tripp1979}, our results also indirectly point toward response planning as a key factor contributing to children's slow responses in real-time interaction \citep{casillasUndRev}.

@@

\bibliographystyle{elsarticle-harv}
\bibliography{anticip}

>>>>

\usepackage{graphicx}
\usepackage{lineno}
\usepackage{amssymb}
\usepackage{wrapfig}
\usepackage{tipa}
\usepackage{float}
\usepackage{setspace}

%\doublespacing

@@

\author[MPI]{Marisa Casillas\corref{cor1}}
\address[MPI]{Max Planck Institute for Psycholinguistics}
%\address[StanfordLX]{Department of Linguistics, Stanford University}
\cortext[cor1]{Corresponding author. Address: Max Planck Institute for Psycholinguistics, Wundtlaan 1, 6525XD, Nijmegen, The Netherlands. \textit{E-mail address:} marisa.casillas@mpi.nl.}
%\newline \newline Address M.C. Frank: Stanford University, Jordan Hall Rm. 420-278, 450 Serra Mall, Stanford, CA 94305. \textit{E-mail address:} mcfrank@stanford.edu

\author[StanfordPSY]{Michael C. Frank}
\address[StanfordPSY]{Department of Psychology, Stanford University}

\begin{abstract}
We investigate language acquisition through the lens of a fundamental conversational skill: turn taking. Children begin developing turn-taking skills in infancy, but take several years to assimilate their growing knowledge of language with this ability. In two eye-tracking experiments with children across a wide developmental range (1;0--6;11), we measured spontaneous predictions about upcoming speaker change while controlling the amount of linguistic information available. We found that children already predicted upcoming turns at age one, but they integrated linguistic cues differently at different ages. Children under three showed a general advantage for prosody over lexicosyntax, contrary to prior findings for adults. Children two and older showed more predictive switches for questions than non-questions, but only when lexical information was available. We found no evidence that lexicosyntax alone guides turn prediction---instead, participants' performance was best overall with access to lexicosyntax and prosody together.
\end{abstract}

@@

Spontaneous conversation is a universal context for using and learning language. Like other types of human interaction (e.g., driving, playing a sport, duetting), it is organized at its core by the roles and goals of its participants. But what sets it apart from other modes of interaction is its structure: sequences of interconnected, communicative acts that take place within alternating turns at talk. Turn-based structures in conversation are strikingly consistent across linguistic communities and  modalities \citep{de-vosInPrep, dingemanse2013, stivers2009}. In particular, the timing of transitions from one speaker to the next is consistently rapid, averaging between 0 and 200 msec. To achieve such rapid turn transitions, addressees must engage in predictive language comprehension and early response planning, motivating the idea that linguistic and interactional processing are deeply entwined in real-time language use, at least for adult speakers \citep{levinson2013}.

Infants and young children manage to participate in their everyday interactions with quite limited linguistic repertoires---children begin taking turns (of a sort) long before their first words emerge \citep[age 3--4 months: see, e.g.,][]{bateson1975, hilbrinkUR, jaffe2001, snow1977} and, once they start using language, it takes a few years before they can give verbal responses with rapid turn transitions \citep{casillas2015, garvey1984, ervin-tripp1979}. Little is known about when and how children's linguistic knowledge begins to influence their  interactional behavior (as it does for adults). One idea is that they start out by using non-linguistic cues to make predictions about turn-taking behavior, and then later integrate linguistic cues as they acquire them. In the current study we investigate when children begin to make predictions about upcoming turn structure, and how they integrate language into their predictions as they grow older.

\subsection{Adults turn prediction}

Prior work suggests that adult listeners primarily use lexicosyntactic information to accurately predict upcoming turn structure \citep{de-ruiter2006}. De Ruiter and colleagues \citeyearpar{de-ruiter2006} asked participants to listen to snippets of spontaneous conversation and to press a button whenever they anticipated that the current speaker was about to finish her turn. The speech snippets were controlled for the amount of linguistic information present; some were normal, but others had flattened pitch (no intonation), low-pass filtered speech (no lexico-syntactic cues), or further manipulations. The timing of participants' button responses with pitch-flattened speech was comparable to their timing with the full linguistic signal but, when no lexical information was available, was significantly earlier than responses for the full linguistic signal. The authors concluded that lexicosyntactic information\footnote{The ``lexicosyntactic'' condition only included flattened pitch and so was not exclusively lexicosyntactic---the speech would still have residual prosodic structure, including syllable duration and intensity.} was necessary and possibly sufficient for turn-end projection, while intonation was neither necessary nor sufficient. Congruent evidence comes from studies varying the predictability of lexicosyntactic and pragmatic content: Adults anticipate turn ends better when they can more accurately predict the exact words that will come next (\citealp{magyari2012}; see also \citealp{magyariUndRev}).

The role of prosody (rhythm, intensity, and intonation) for adult turn-prediction is still a matter of debate. Prosodic structure is closely tied to the syntax of an utterance, making the two linguistic signals difficult to control independently \citep{ford1996}. \citet*{torreiraUndRev} measured the turn-end predictive effects of prosody and syntax by splicing together separate segments of speech to create syntactically-complete stimuli with and without intonational phrase boundaries at the end. In their study, participants never responded (verbally or with a ``turn-end'' button-press)  without hearing an intonational phrase boundary. On the other hand, when they embedded intonational phrase boundaries in multi-utterance turns, participants were tricked into pressing the ``turn-end'' button on 29\% of trials. Listeners then \textit{do} rely on prosodic cues to predict upcoming turn structure (see also \citet{de-ruiter2006}:525), perhaps using them in combination with other linguistic cues (lexicosyntactic, and pragmatic) for greater accuracy in response timing \citep{duncan1972, ford1996, levinson2013, hirvenkari2013}.

Taken together, these corpus-based and experimental findings on human conversation suggest that linguistic cues may be critical to smooth turn transitions in adult conversation. If so, we need an account of how children learn to integrate linguistic cues into their interactive turn taking over the course of language acquisition.

\subsection{Child turn prediction}

@@

Children's first turn-like structures appear as early as two to three months after birth, in proto-conversation with their caregivers \citep{bruner1975, bruner1985}. During proto-conversations, caregivers interact with their infants as if they were capable of making meaningful contributions; they take every look, vocalization, arm flail, and burp as ``utterances'' in the joint discourse \citep{bateson1975, jaffe2001, snow1977}. Infants catch onto the structure of proto-conversations quickly. By three to four months they notice disturbances to the contingency of caregiver responses and, in reaction, change the rate and quality of their vocalizations \citep{k-bloom1988, masataka1993}.

The timing of children's responses to their caregivers' speech shows a non-linear pattern from early infancy to middle childhood. Infants' turn timing at three months is often too early or too late; they overlap their vocalizations with caregiver speech nearly 40\% of the time and come in after inter-turn silent gaps that average around 675 msec (adults: 200 msec; \citealp{hilbrinkUR}). Between four and nine months children reduce the number of turns happening in overlap. At the same time, children's silent gaps temporarily begin to get shorter, before then getting longer again, peaking at an average duration of 1100 msec at nine months. After nine months, children's gap durations continue getting shorter again, but very gradually \citep{casillas2015, hilbrinkUR}; while children's avoidance of overlap is nearly adult-like by nine months, the timing of their non-overlapped responses stays much longer than the 200 msec standard for the next several years \citep{garvey1984, ervin-tripp1979}.

The gradual development of children's turn timing after age nine months may be attributable to linguistic development; taking turns on time is easier when the response is a simple vocalization rather than a linguistic utterance. If response planning (i.e., language production) is the primary hurdle in children's spontaneous turn taking, children might understand turn-taking behaviors before they are able to produce the behaviors themselves; this hypothesis has been explored recently in experimental contexts.

@@

Children develop specific expectations about conversational behavior before they begin to speak. For example, sometime between four and six months, children begin attending differently to face-to-face and back-to-back conversation; six-month-olds follow conversational speakers more with their gaze when at least one speaker is looking at the other \citep{augusti2010}. At ten months, infants expect people to look and talk at other people, and not to objects \citep{beier2012}.

There are mixed results regarding when children begin to anticipate turn structure in conversation. One study found that 12-month-olds make more predictive gaze shifts to a responder while watching human verbal conversation compared to ``conversation'' between two objects \citep{bakker2011}, but another study only found a similar effect at 36 months \citep{hofsten2009}. Neither study compared children's looking behavior to a random baseline, even though baseline estimates can help account for non-turn-contingent developmental differences in gaze shifting. A third study, testing turn-transition prediction with 6-, 12-, 24-, and 36-month-olds (and adult controls), found that children's rate of anticipatory gaze switching to the responder was only greater than chance for 36-month-olds and adults \citep{keitel2013}. In their study, participants saw two types of conversation videos: one with pitch-flattened speech and one with normal speech. The 36-month-olds, but not the adults, were affected by a lack of intonation contours, leading the authors to conclude that children's ability to predict upcoming turn structure relies on their ability to retrieve lexicosemantic meaning from the stimuli (emerging around 36 months). They suggest that, once children acquire sophisticated, adult-like language comprehension (sometime after 36 months),  intonation takes a secondary role in turn prediction (but also see \citealp{keitel2015}), consistent with other findings on adult turn-end prediction \citep{ruiter2006}.

An early role for intonation is sensible given that children begin acquiring prosodic knowledge early in infancy. Children can distinguish their native language's rhythm type from others (e.g., syllable-timed vs. stress-timed) 2--5 days after birth \citep{mehler1988, moon1993, nazzi2003} and, before 8 months, they can already identify prosodic boundaries and use them to parse information in the speech stream \citep{johnson2001, jusczyk1993, jusczyk1995, morgan1995, soderstrom2003}. In comparison, 8-month-olds still have quite small lexical inventories and, even at 12--14 months, children have just begun to work out syntactic categories \citep{bergelson2013, shi2010}. Two-month-olds notice changes in word order, but this ability appears to rely on prosodic cueing \citep{mandel1996}. Generally speaking then, we might predict from children's linguistic development that children might use prosodic cues in their turn-taking predictions before they use lexicosyntactic cues. Keitel et al.'s \citeyearpar{keitel2013, keitel2015} findings support this idea, but suggest that children's ability to make predictions about upcoming turn structure only emerges at 36 months.

Although the Keitel et al. \citeyearpar{keitel2013, keitel2015} studies constitutes a substantial advance over previous work, it has several limitations. First, the estimates of ``random'' baseline shift frequency were computed exclusively from switches that occurred during the speaker's turn whereas estimates of anticipatory shifting frequency were computed exclusively from switches that occurred at points of speaker switch. Assuming that gaze switches happen most often during at points of speaker switch \citep{hirvenkari2013}, this analysis maximizes the probability of finding a difference between ``random'' and anticipatory switch frequencies. A truly random baseline would instead compare switches at turn transitions to switches at randomly-selected time points throughout the stimuli. Second, the conversation stimuli were somewhat unusual: the inter-turn gaps were longer than is typical for conversation ($~$900 msec), the conversations were scripted and adult-directed, and the speakers were recorded and played back on the same audio channel, making the two voices harder for viewers to distinguish. Despite these methodological issues, the studies demonstrate intriguing age-based differences in children's ability to predict upcoming turn structure. The results suggest that both semantic and intonational development \textit{do} play a role in children's looking patterns. Our current work takes this paradigm as our starting point.\footnote{See also \citet{casillas2012, casillas2013}.}

\subsection{The Current Study}
We report on the role of linguistic processing in children's predictions about upcoming turn structure. In two eye-tracking experiments, we measured children's anticipatory gaze to upcoming responders while controlling for the amount of lexicosyntactic and prosodic information available. In Experiment 1 we showed English-speaking 3--5-year-olds and adults conversation video clips in several different languages to control for the presence of lexicosyntactic information while leaving prosodic and non-linguistic cues intact. This paradigm showed minimal differences between the predictive looking behavior of preschoolers and adults. In Experiment 2, we used artificial (puppet) visual stimuli and acoustic manipulation in English-only videos to control the presence of prosodic and lexical cues in the stimuli while eliminating predictive visual cues to turn taking (e.g., gaze and gesture). In this more controlled paradigm, we found that children's predictive looking behavior improved from ages one to six, but that even one-year-olds made more anticipatory looks than would be expected by chance.

%In both experiments children consistently looked faster to responders after hearing questions, compared to non-questions. Both prosodic and lexicosyntactic information played a role in children's predictions about turn structure, but the two information sources were used differently at different ages. Our findings generally support an account in which predictive processes for turn taking in conversation are present early, but their integration with linguistic information is gradual.

@@

\section{General Discussion}
\label{sec:gendisc}
Children begin to develop conversational turn-taking skills long before their first words \citep{bateson1975, hilbrinkUR, jaffe2001, snow1977}. As they acquire language, they also acquire the information needed to make accurate predictions about upcoming turn structure. Until recently, we have had very little data on how children weave language into their already-existing turn-taking behaviors.

@@

Our findings are consistent with observational work on children's turn taking. Young children are slow turn-takers in spontaneous conversation \citep{casillas2015} even though they begin taking turns in infancy \citep{hilbrinkUR, jaffe2001}. We found that children begin to spontaneously predict upcoming turn structure at age one. Children's gaze switches were also most common in the inter-turn gap, suggesting that they expected an immediate response from the addressee. Our results therefore support the idea of an early-developed competence for the timing of turn taking. But because children do not themselves use adult-like conversational timing until much later \citep[age six; ][]{ervin-tripp1979}, our results also indirectly point toward response planning as a key factor contributing to children's slow responses in real-time interaction \citep{casillas2015}.

@@

\bibliographystyle{elsarticle-harv}
\bibliography{anticip-Edits}
